---
title: "Konza Vegetation Analysis"
author: "Luke Hackler"
format: html
---

## Introduction

This analysis will look at vegetation data from the Konza National Ecological Observatory Network (NEON) site located near Manhattan, KS. It will use tidyverse packages to wrangle and visualize some of the data.

## Loading the data

The data are provided in csv files. We'll use the tidyverse package `readr` to load it into R.

```{r}
#| label: read-csv

# load packages
library("readr")
library("dplyr")
library("ggplot2")

# read in the data
konza_data <- read_csv("data/NEON.D06.KONZ.DP1.10058.001.div_1m2Data.2025-08.basic.20251222T234053Z.csv")
```

## Data summarizing with `dplyer` 

We'll next use the dplyer package to summarize some different aspects of the data.

```{r}
#| label: summarize-with-dplyer

# first summary to make will be a count of the number of observations in each family
konza_data %>%
  group_by(family) %>%
  tally() %>%
  arrange(desc(n))
```

This analysis showed that Poaceae and Asteraceae are among the most common family found in these plots.

```{r}
#| label: mean-percent-cover-by-family

# the goal is to calculate the mean percent cover of each family
konza_data %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover))
```

This analysis showed that the largest percent cover is the `NA` taxa, which means it likely wasn't able to be identified, but the highest mean percent cover at 16% across the plots was Fagaceae.

```{r}
#| label: filter-by-elevation

# the goal is to calculate the mean percent cover of each family
konza_data %>%
  filter(elevation > 390) %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover))
```

Next we'll learn to use `mutate()`, which is a function to create a new column based on contents from an existing column

```{r}
#| label: use-mutate

# select pulls out the columns you want
# mutate creates a new column with whatever content you want
# which can include manipulation of an existing column
konza_data %>%
  select(scientificName, family, percentCover) %>%
  mutate(propCover = percentCover / 100)
```

Next up, we'll use `ggplot2` to make some figures!

```{r}
#| label: elevation-percent-by-family

konza_data %>%
  filter(!is.na(family)) %>%
  filter(family == "Asteraceae") %>%
  ggplot() +
  geom_boxplot(aes(x = elevation,
                   y = percentCover,
                   group = elevation)) +
  labs(title = "Cover of Asteraceae by Elevation",
       y = "Percent Cover",
       x = "Elevation")
```


## Next steps

Need to demonstrate the use of the `mutate()` function, and then how to use `ggplot2`. Some interesting plots could be looking at the effect of elevation on plant cover.


## Graph 1: I want to see how percent cover varies between plot types 
## (tower and distributed)

```{r}
#| label: plot-type-and-percent-cover

konza_data %>%
  filter(plotType == "tower" | plotType == "distributed") %>%
  ggplot() +
  aes(x = plotType,
      y = percentCover) +
  geom_boxplot() +
  labs(title = "Percent Cover by Plot Type",
       x = "Plot Type",
       y = "Percent Cover")
```

## Glad that the plot types are appropriately named. The distributed plot type
## shows 

## Graph 2: Next I will try to visualize the counts of Poaceae in
## each 'nlcdClass'

```{r}
#| label: asteraceae-count-by-nlcd-class

konza_data %>%
  filter(family == "Poaceae") %>%
  ggplot() +
  aes(x = nlcdClass,
      fill = family) +
  geom_bar() +
  scale_fill_manual(values = c("forestgreen")) +
  labs(title = "Poaceae per NLCD Class",
       x = "NLCD Class",
       y = "Poaceae Count")
```

## I used the '?geom_bar' command to get help using `geom_bar`. The bar graph
## showed that more Poaceae counts occurred in Grassland Herbaceous

## Graph 3: For my final graph I want to make a graph using `geom_point`
## and `geom_smooth`. I also want to use `mutate` to get the proportion
## cover of specific families and compare them to their elevations. I 
## want to use `filter` to filter proportion cover below 0.15 and elevation
## above 380. Also add fun colors using `scale_color_manual`!

```{r}
#| label: prop-cover-poa-aster-corn-by-elevation

konza_data %>%
  select(family, elevation, percentCover) %>%
  mutate(propCover = percentCover / 100) %>%
  filter(family == "Poaceae" | family == "Asteraceae" | family == "Cornaceae") %>%
  filter(propCover < 0.15) %>%
  filter(elevation > 380) %>%
  ggplot() +
  aes(x = elevation,
      y = propCover,
      color = family) +
  geom_point() +
  geom_smooth() +
  scale_color_manual(values = c("Poaceae" = "hotpink",
                                "Asteraceae" = "limegreen",
                                "Cornaceae" = "purple")) +
  labs(title = "Proportion Cover of Poaceae, Asteraceae, and Cornaceae by Elevation",
       x = "Elevation (m)",
       y = "Proportion Cover")
```

## This figure shows Poaceae declining in proportion cover as elevation
## increases, Asteraceae rising in proportion cover as elevation increases
## and Cornaceae holding a constant line supported by data points that
## seem to be evening eachother out. 
